---
title: "mediation"
author: "Dong Wang"
date: '2022-06-02'
output:
  html_document: default
  pdf_document: default
---

# Setup
## Load Libraries (install them if needed)
```{r}
library("readxl")
library(naniar)
library(MASS)
library(Matrix)
library(mvtnorm)
library(sandwich)
library(mediation)
source('MedOmniTest.R')
source('MedTest-Internal.R')
source('modima.R')
source('microhima.R')
library('matrixStats')
library('vegan')
library('GUniFrac')
library('ape')
library(data.table)
library(MultiMed)
library(energy)
```

## Load all data here, including feature table, metadata, phylogenetic tree
```{r}
otu.tab <- fread("row-normalized-table.csv", header=TRUE)
otu.tab = subset(otu.tab, select = -c(diagnosis))
family.tab <- fread("family-table.csv", header=TRUE)
genus.tab <- fread("collapsed-table-genus.csv", header=TRUE)
gut.meta <- read.csv("metadata.tsv", sep="\t", header=TRUE)
tree.rooted <- read.tree(file='insertion_tree.relabelled.tre')

# Process feature table (row and column names)
otu.tab = data.frame(otu.tab)
otu.tab <- data.frame(otu.tab[,-1], row.names = otu.tab[,1])
family.tab = data.frame(family.tab)
family.tab <- data.frame(family.tab, row.names = family.tab[,1])
genus.tab = data.frame(genus.tab)
genus.tab <- data.frame(genus.tab, row.names = genus.tab[,1])
```


## Process metadata (get features of interest)
```{r}
# Process metadata (get features of interest)
Result = data.frame(gut.meta)
Result = subset(Result, select = c(sampleid, diagnosis, depression_cesd10_score, anxiety_score, cognitive_failures_score, years_smoker, bmi, sys, age_corrected, smoking_habits, sex, race_cat, antibiotic_use) )
Result$diagnosis<-ifelse(Result$diagnosis=="SZ",1,0)
Result$smoking_habits<-ifelse(Result$smoking_habits=="Yes",1,0)
Result$antibiotic_use<-ifelse(Result$antibiotic_use=="Yes",1,0)
Result$sex<-ifelse(Result$sex=="male",1,0)
Result$race_cat<-ifelse(Result$race_cat=="Caucasian",1,0)
```


# Mediation Analysis with MedTest
## Smoker variable
```{r}
# smoker
unifracs <- GUniFrac(otu.tab, tree.rooted)$unifracs
m.list <- list(BC=vegdist(otu.tab, method="bray"),
		JAC=as.matrix(vegdist(otu.tab, 'jaccard', binary=TRUE)),
		# UniFrac=unifracs[, , c('d_UW')],
		GUniFrac=unifracs[, , c('d_0.5')])
		# WUniFrac=unifracs[, , c('d_1')])

set.seed(12345)
MedOmniTest(Result$smoking_habits, Result$diagnosis, m.list)
```


## bmi variable
```{r}
# bmi
new_frame = otu.tab
new_frame$exposure = Result$bmi # change this here for each exposure variable
new_frame$diagnosis = Result$diagnosis
new_frame = new_frame[!is.na(as.numeric(new_frame$exposure)), ] 
exposure = as.numeric(new_frame$exposure)
diagnosis = as.numeric(new_frame$diagnosis)
new_frame = subset(new_frame, select = -c(exposure, diagnosis))

unifracs <- GUniFrac(new_frame, tree.rooted)$unifracs
m.list <- list(BC=vegdist(new_frame, method="bray"),
		JAC=as.matrix(vegdist(new_frame, 'jaccard', binary=TRUE)),
		# UniFrac=unifracs[, , c('d_UW')],
		GUniFrac=unifracs[, , c('d_0.5')])
		# WUniFrac=unifracs[, , c('d_1')])
set.seed(12345)
MedOmniTest(exposure, diagnosis, m.list)
# ModimaTest <- modima(exposure, new_frame, diagnosis)
# ModimaTest$estimates
```


## years smoker variable
```{r}
# years smoker
new_frame = otu.tab
new_frame$exposure = Result$years_smoker
new_frame$diagnosis = Result$diagnosis
new_frame = new_frame[!is.na(as.numeric(new_frame$exposure)), ] 
exposure = as.numeric(new_frame$exposure)
diagnosis = as.numeric(new_frame$diagnosis)
new_frame = subset(new_frame, select = -c(exposure, diagnosis))

unifracs <- GUniFrac(new_frame, tree.rooted)$unifracs
m.list <- list(BC=vegdist(new_frame, method="bray"),
		JAC=as.matrix(vegdist(new_frame, 'jaccard', binary=TRUE)),
		# UniFrac=unifracs[, , c('d_UW')],
		GUniFrac=unifracs[, , c('d_0.5')])
		# WUniFrac=unifracs[, , c('d_1')])
set.seed(12345)
MedOmniTest(exposure, diagnosis, m.list)
```


## anxiety score variable
```{r}
# anxiety score
new_frame = otu.tab
new_frame$exposure = Result$anxiety_score
new_frame$diagnosis = Result$diagnosis
new_frame = new_frame[!is.na(as.numeric(new_frame$exposure)), ] 
exposure = as.numeric(new_frame$exposure)
diagnosis = as.numeric(new_frame$diagnosis)
new_frame = subset(new_frame, select = -c(exposure, diagnosis))

unifracs <- GUniFrac(new_frame, tree.rooted)$unifracs
m.list <- list(BC=vegdist(new_frame, method="bray"),
		JAC=as.matrix(vegdist(new_frame, 'jaccard', binary=TRUE)),
		# UniFrac=unifracs[, , c('d_UW')],
		GUniFrac=unifracs[, , c('d_0.5')])
		# WUniFrac=unifracs[, , c('d_1')])
set.seed(12345)
MedOmniTest(exposure, diagnosis, m.list)
```


## depression score variable
```{r}
# depression score
new_frame = otu.tab
new_frame$exposure = Result$depression_cesd10_score
new_frame$diagnosis = Result$diagnosis
new_frame = new_frame[!is.na(as.numeric(new_frame$exposure)), ] 
exposure = as.numeric(new_frame$exposure)
diagnosis = as.numeric(new_frame$diagnosis)
new_frame = subset(new_frame, select = -c(exposure, diagnosis))

unifracs <- GUniFrac(new_frame, tree.rooted)$unifracs
m.list <- list(BC=vegdist(new_frame, method="bray"),
		JAC=as.matrix(vegdist(new_frame, 'jaccard', binary=TRUE)),
		# UniFrac=unifracs[, , c('d_UW')],
		GUniFrac=unifracs[, , c('d_0.5')])
		# WUniFrac=unifracs[, , c('d_1')])
set.seed(12345)
MedOmniTest(exposure, diagnosis, m.list)
```


## age variable
```{r}
# age
new_frame = otu.tab
new_frame$exposure = Result$age_corrected
new_frame$diagnosis = Result$diagnosis
new_frame = new_frame[!is.na(as.numeric(new_frame$exposure)), ] 
exposure = as.numeric(new_frame$exposure)
diagnosis = as.numeric(new_frame$diagnosis)
new_frame = subset(new_frame, select = -c(exposure, diagnosis))

unifracs <- GUniFrac(new_frame, tree.rooted)$unifracs
m.list <- list(BC=vegdist(new_frame, method="bray"),
		JAC=as.matrix(vegdist(new_frame, 'jaccard', binary=TRUE)),
		# UniFrac=unifracs[, , c('d_UW')],
		GUniFrac=unifracs[, , c('d_0.5')])
		# WUniFrac=unifracs[, , c('d_1')])
set.seed(12345)
MedOmniTest(exposure, diagnosis, m.list)
```


## sex variable
```{r}
# sex
new_frame = otu.tab
new_frame$exposure = Result$sex
new_frame$diagnosis = Result$diagnosis
new_frame = new_frame[!is.na(as.numeric(new_frame$exposure)), ] 
exposure = as.numeric(new_frame$exposure)
diagnosis = as.numeric(new_frame$diagnosis)
new_frame = subset(new_frame, select = -c(exposure, diagnosis))

unifracs <- GUniFrac(new_frame, tree.rooted)$unifracs
m.list <- list(BC=vegdist(new_frame, method="bray"),
		JAC=as.matrix(vegdist(new_frame, 'jaccard', binary=TRUE)),
		# UniFrac=unifracs[, , c('d_UW')],
		GUniFrac=unifracs[, , c('d_0.5')])
		# WUniFrac=unifracs[, , c('d_1')])
set.seed(12345)
MedOmniTest(exposure, diagnosis, m.list)
```


## race variable
```{r}
# race
new_frame = otu.tab
new_frame$exposure = Result$race_cat
new_frame$diagnosis = Result$diagnosis
new_frame = new_frame[!is.na(as.numeric(new_frame$exposure)), ] 
exposure = as.numeric(new_frame$exposure)
diagnosis = as.numeric(new_frame$diagnosis)
new_frame = subset(new_frame, select = -c(exposure, diagnosis))

unifracs <- GUniFrac(new_frame, tree.rooted)$unifracs
m.list <- list(BC=vegdist(new_frame, method="bray"),
		JAC=as.matrix(vegdist(new_frame, 'jaccard', binary=TRUE)),
		# UniFrac=unifracs[, , c('d_UW')],
		GUniFrac=unifracs[, , c('d_0.5')])
		# WUniFrac=unifracs[, , c('d_1')])
set.seed(12345)
MedOmniTest(exposure, diagnosis, m.list)
```


## antibiotic variable
```{r}
# antibiotic
new_frame = otu.tab
new_frame$exposure = Result$antibiotic_use
new_frame$diagnosis = Result$diagnosis
new_frame = new_frame[!is.na(as.numeric(new_frame$exposure)), ] 
exposure = as.numeric(new_frame$exposure)
diagnosis = as.numeric(new_frame$diagnosis)
new_frame = subset(new_frame, select = -c(exposure, diagnosis))

unifracs <- GUniFrac(new_frame, tree.rooted)$unifracs
m.list <- list(BC=vegdist(new_frame, method="bray"),
		JAC=as.matrix(vegdist(new_frame, 'jaccard', binary=TRUE)),
		# UniFrac=unifracs[, , c('d_UW')],
		GUniFrac=unifracs[, , c('d_0.5')])
		# WUniFrac=unifracs[, , c('d_1')])
set.seed(12345)
MedOmniTest(exposure, diagnosis, m.list)
```



<!-- # Experimental teseting with MultiMed's medTest -->
<!-- This tests for multiple biomarkers as mediators simultaneously.  -->
<!-- No significant species/genus/family was found in this dataset -->

<!-- ## smoker variable -->
<!-- ```{r} -->
<!-- medSmoker <- medTest(E=Result$smoking_habits, M=family.tab, Y=Result$diagnosis, nperm = 9999) -->
<!-- rownames(medSmoker) <- colnames(family.tab) -->
<!-- medSmoker[which.min(medSmoker[,"p"]),,drop=FALSE] -->
<!-- medSmoker[,"p"] -->
<!-- ``` -->

<!-- ## anxiety score variable -->
<!-- ```{r} -->
<!-- new_frame = genus.tab -->
<!-- new_frame$exposure = Result$anxiety_score -->
<!-- new_frame$diagnosis = Result$diagnosis -->
<!-- new_frame = new_frame[!is.na(as.numeric(new_frame$exposure)), ]  -->
<!-- exposure = as.numeric(new_frame$exposure) -->
<!-- diagnosis = as.numeric(new_frame$diagnosis) -->
<!-- new_frame = subset(new_frame, select = -c(exposure, diagnosis)) -->
<!-- print(diagnosis) -->

<!-- medSmoker <- medTest(E=exposure, M=new_frame, Y=diagnosis, nperm = 9999) -->
<!-- rownames(medSmoker) <- colnames(genus.tab) -->
<!-- medSmoker[which.min(medSmoker[,"p"]),,drop=FALSE] -->
<!-- medSmoker[,"p"] -->
<!-- ``` -->

<!-- ```{r} -->
<!-- medSmoker <- medTest(E=Result$smoking_habits, M=otu.tab, Y=Result$diagnosis, nperm = 999) -->
<!-- rownames(medSmoker) <- colnames(otu.tab) -->
<!-- medSmoker[which.min(medSmoker[,"p"]),,drop=FALSE] -->
<!-- # medSmoker[,"p"] -->
<!-- ``` -->

